---
title: "Google Capstone Project_Cyclistic Bike-Share"
author: "Damber Singh Biswa"
date: "2023-02-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparing data for analysis ##
**Installing libraries in R **
```{r}
# tidyverse for data wrangling, lubridate for formatting date and ggplot2 for visualization.
library(tidyverse) 
library(lubridate) 
library(ggplot2)
library(dplyr)
library(janitor)
library(readr) # used for reading a csv file.
```
```{r}
# Displaying and setting-up working directory
getwd()
setwd("C:\\Users\\dambe\\Desktop\\Portfolio Project")
```
```{r}
# Loading dataset for combining
m1<-read.csv("1_March_2022.csv", na="")
m2<-read.csv("2_April_2022.csv",na="")
m3<-read.csv("3_May_2022.csv",na="")
m4<-read.csv("4_June_2022.csv",na="")
m5<-read.csv("5_July_2022.csv",na="")
m6<-read.csv("6_August_2022.csv",na="")
m7<-read.csv("7_Sept_2022.csv",na="")
m8<-read.csv("8_Oct_2022.csv",na="")
m9<-read.csv("9_Nov_2022.csv",na="")
m10<-read.csv("10_Dec_2022.csv",na="")
m11<-read.csv("11_Jan_2023.csv",na="")
m12<-read.csv("12_Feb_2023.csv",na="")
```

__Comparing column names of each files.__
```{r}
colnames(m1)
colnames(m2)
colnames(m3)
colnames(m4)
colnames(m5)
colnames(m6)
colnames(m7)
colnames(m8)
colnames(m9)
colnames(m10)
colnames(m11)
colnames(m12)
```

** combining multiple data sets to a single large data set.**
```{r}
combined_trip <- bind_rows(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12);
View(combined_trip)
```
## Processing data for analysis ##
**Evaluating data for integrity**
```{r}
head(combined_trip)
nrow(combined_trip)
ncol(combined_trip)
glimpse(combined_trip)
```


**Dropping any rows with NA/empty values**
```{r}
combined_trip2 <- remove_empty(combined_trip, c("rows", "cols"))%>%
  drop_na()
```
**Inspecting the the cleaned data**
```{r}
glimpse(combined_trip2)
str(combined_trip2)
```

**Adding columns date, month, day, year and hour of each ride**
```{r}
# https://www.statmethods.net/input/dates.html more on date formats in R found at that link
combined_trip2$Date <- as.Date(combined_trip2$started_at)
combined_trip2$Month <- format(as.Date(combined_trip2$Date), "%m")
combined_trip2$Day <- format(as.Date(combined_trip2$Date), "%d")
combined_trip2$Year <- format(as.Date(combined_trip2$Date), "%Y")
combined_trip2$Day_of_Week <- format(as.Date(combined_trip2$Date), "%A")
combined_trip2$Hour <- hour(combined_trip2$started_at)
combined_trip2$Ride_Length <- difftime(combined_trip2$ended_at,combined_trip2$started_at,units = "min") #calculating ride_length in minutes
```

**Inspecting additional columns and structure of column**
```{r}
str(combined_trip2)
```
#removing unwanted columns from the data set
```{r}
combined_trip2 <- subset(combined_trip2, select = -c(started_at, ended_at,start_station_id, end_station_id,start_lat, start_lng,end_lat, end_lng))
str(combined_trip2)
```

**Convert "ride_length" from factor to numeric for easy calculations**
```{r}
combined_trip2$Ride_Length <- as.numeric(as.character(combined_trip2$Ride_Length))
is.numeric(combined_trip2$Ride_Length)
str(combined_trip2)
```

**Inspecting and removing bad data **
```{r}
#Checking ride length less than 0
nrow(subset(combined_trip2, Ride_Length<0))

# Removing row with ride_length less than 0.
combined_trip2 <- combined_trip2[!(combined_trip2$Ride_Length<0),]

#Checking if any bicycles were used for test/by company for maintenance
nrow(subset(combined_trip2,start_station_name == "HQ QR"))
```

## Descriptive Analysis ##
*Descriptive analysis on ride_length (all figures in minutes) *
```{r}
mean(combined_trip2$Ride_Length)
median(combined_trip2$Ride_Length)  #median ride length
max(combined_trip2$Ride_Length) #max ride length
min(combined_trip2$Ride_Length) # shortest ride length
```

**Compare members and casual users by aggregate**
```{r}
aggregate(combined_trip2$Ride_Length ~ combined_trip2$member_casual, FUN = mean)
aggregate(combined_trip2$Ride_Length ~ combined_trip2$member_casual, FUN = median)
aggregate(combined_trip2$Ride_Length ~ combined_trip2$member_casual, FUN = max)
aggregate(combined_trip2$Ride_Length ~ combined_trip2$member_casual, FUN = min)
```
**ordering days of week in sequence ** 
```{r}
combined_trip2$Day_of_Week <- ordered(combined_trip2$Day_of_Week, 
                                       levels=c("Sunday", "Monday", "Tuesday", 
                                                "Wednesday", "Thursday", "Friday", 
                                                "Saturday")) 
```


**Average ride time by each day for members vs casual users**

```{r}
aggregate(combined_trip2$Ride_Length ~ combined_trip2$member_casual + combined_trip2$Day_of_Week, FUN = mean)
```

## Sharing /Visualization ##
**Analyze ridership data by membership type and weekday**
```{r}
combined_trip2 %>%  
  group_by(member_casual, Day_of_Week) %>% 
  summarise(avg_trip_time_in_min = mean(Ride_Length)) %>%
  ggplot(aes(x = Day_of_Week, y = avg_trip_time_in_min, fill = member_casual)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average trip duration by customer type Vs. Day of the week")
```

**Number of ride by day of week**
```{r}
combined_trip2 %>% 
  group_by(rideable_type, Day_of_Week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(Ride_Length)) %>% 
  arrange(rideable_type, Day_of_Week)  %>% 
  ggplot(aes(x = Day_of_Week, y = number_of_rides, fill = rideable_type)) +
  geom_col(position = "dodge") + 
  labs(title = "Number of Rides by Day and Bike Type") + 
  ylab("Number of Rides") + 
  xlab("Day of Week")+geom_text(aes(label=number_of_rides))
```


## Exporting cleaned data set for further analysis##
```{r}
write.csv(combined_trip2,"Final_Cleaned_Trip_Data.csv", row.names = TRUE)

``` 
